### 2015-06-06 chapters/ipa_meets_unicode.tex
finish pitfall IPA homoglyphs

\section{Pitfall: Homoglyphs in IPA}

It is not just the Unicode Standard that offers multiple options for encoding 
the IPA. Even the IPA itself offers some flexibility in how specific meaningful 
transcriptions have to be encoded. There are a few cases in which the IPA 
explicitly allows for different options of transcribing the same phonetic 
content. This is understandable from a transcriber's point of view, but it is 
not good interoperability between resources supposedly written in IPA. We 
consider it crucial to distinguish between ``lax'' IPA, in which the 
phonetically-trained reader should be able to understand the transcription, and 
``strict'' IPA, which should be standardized on a single encoding so search will 
work across resources.

Another example we commonly encounter is the use of <g> \textsc{latin small
letter g} at \uni{0067}, instead of the Unicode Standard IPA character for the
voiced velar stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One
begins to question whether this issue is at all apparent to the working
linguist, or if they simply use the \uni{0067} because it is easily keyboarded
and thus saves time, whereas the latter must be cumbersomely inserted as a
special symbol in most software.\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}

\begin{itemize}
  \item tone marks (prefer tone symbols)
  \item marking of <g>  \item marking of voiceless with ring diacritic
  \item tie bar
\end{itemize}

@@

One important distinction to acknowledge is the difference between multigraphs
and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
< tʃ > or < ou >) while ligatures are single characters (e.g. <ʧ> \textsc{latin
small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
printing easier-to-read texts, and are included in the Unicode Standard for
reasons of legacy encoding. However, their usage is discouraged by the Unicode
core specification. Specifically related to IPA, various phonetic combinations
of characters (typically affricates) are available as single code-points in the
Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
(confusingly both names appear interchangeably). Such glyphs might be used by
software to produce a pleasing display, but they should not be hard-coded into
the text itself. In the context of IPA, characters like the following ligatures
should thus \emph{not} be used. Instead a combination of two characters is
preferred:

>>>>

\section{Pitfall: Homoglyphs in IPA}
\label{pitfall-homoglyphs-in-IPA}

It is not just the Unicode Standard that offers multiple options for encoding
the IPA. Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA. We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources.

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. This issue was recently addressed by The International Phonetic
Association has taken the stance that both the keyboard \textsc{latin small
letter g} and the \textsc{latin small letter script g} are valid input
characters for the voiced velar plosive. Unfortunately, this decision further
introduces ambiguity for linguists trying to adhere to a strict Unicode Standard
IPA encoding. For strict IPA encoding we propose to standardize on the more 
idiosyncratic \textsc{latin small letter script g} at \uni{0261}.

Third, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below.

Likewise, the tie bar to indicate affricates, doubly articulated consonants or
diphthongs can be placed either below or on top of the base characters, e.g.
<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
above the base characters, using Unicode \textsc{combining double inverted
breve} at \uni{0361}. Note that this Unicode character is placed between the two 
base characters to be combined. In principle, it is possible to combine more 
than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
necessary, we consider this possible, even though the rendering will never look 
good.

@@

One important distinction to acknowledge is the difference between multigraphs
and ligatures. Multigraphs are groups of characters (in the context of IPA e.g.
<tʃ> or <ou>) while ligatures are single characters (e.g. <ʧ> \textsc{latin
small letter tesh digraph} at \uni{02A7}). Ligatures arose in the context of
printing easier-to-read texts, and are included in the Unicode Standard for
reasons of legacy encoding. However, their usage is discouraged by the Unicode
core specification. Specifically related to IPA, various phonetic combinations
of characters (typically affricates) are available as single code-points in the
Unicode Standard, but are designated as \textsc{ligatures} or \textsc{digraphs}
(confusingly both names appear interchangeably). Such glyphs might be used by
software to produce a pleasing display, but they should not be hard-coded into
the text itself. In the context of IPA, characters like the following ligatures
should thus \emph{not} be used. Instead a combination of two characters is
preferred:
